Investigation of thermal reactionomes reveals stress tolerance vs. resistance in

### 2014-09-22 doc/ApTranscriptome_manuscript.Rmd
writing

To gain insight on the genetic basis of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions. First, species from more thermally-stressful environments may rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. Finding that fewer genes are up-regulated to high temperatures in the warmer-climate species *A. carolinensis* would support this prediction. Second, spatially-divergent selection often results in local adaptation [@turesson1922]. For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature. Evidence for local adaptation in gene expression would be observed if heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea*, and conversely, that cold-responsive genes are activated at colder temperatures in *A. picea*.

In addition, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

Importantly, none of the above predictions depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012; @pavlidis2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].

To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

## Species comparisons in the number of thermally-responsive genes

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes. The **High** genes in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at **Low** temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina ( 36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

>>>>

To gain insight on the genetic basis of thermal adaptation, we compared the thermal reactionomes of these species with two specific predictions. First, species from more thermally-stressful environments may rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. Finding that fewer genes are up-regulated to high temperatures in the warmer-climate species *A. carolinensis* would support this prediction. Second, spatially-divergent local adaptation leaves a signal at the molecular level [@savolainen2013]. For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature. Evidence for local adaptation in gene expression would be observed if heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea*, and conversely, that cold-responsive genes are activated at colder temperatures in *A. picea*.

In addition, we used the thermal reactionomes to examine the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. Specifically, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction. While none of the above predictions depend on the functional annotation of the responsive genes [@pavlidis2012], insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011]. To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

## Species differences in thermally-responsive genes

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the mean temperatures of gene activation, defined as the temperatures at which the gene showed the greatest change in expression, for genes turned on at high or low temperatures. Genes turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at low temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.
